Separating Use and Reuse to Improve Both

### 2017-05-29 FACS/Examples.tex
Local changes

\saveSpace\saveSpace\section{Intuitions on formalization}\label{sec:formal}
\saveSpace\saveSpace
In this article we dedicate more space to examples and informal presentation and motivations;
so we do not have space to provide a full formalizations.
We will provide here some hints on how the formalization works.

@@

\paragraph{Well-formedness}
Basic well formedness rules apply:
\begin{itemize}
\item all method parameters have unique names and the special parameter name \Q@this@ is not declared
 in the parameter list,
\item all methods in a code literal have unique names,
\item all used variables are in scope,
\item all methods in an interface are abstract, and there are no interface class methods.
\end{itemize}
Those rules can be applied on any given \mL~individually and in full isolation.

>>>>

\saveSpace\saveSpace\section{Intuitions on formalization}\label{sec:formal}
\saveSpace\saveSpace

This section sketches some of the key ideas in \name.

%In this article we dedicate more space to examples and informal presentation and motivations;
%so we do not have space to provide a full formalizations.
%We will provide here some hints on how the formalization works.

\subsection{Syntax}

@@

\subsection{Well-formedness}
Basic well formedness rules apply:
\begin{itemize}
\item all method parameters have unique names and the special parameter name \Q@this@ is not declared
 in the parameter list,
\item all methods in a code literal have unique names,
\item all used variables are in scope,
\item all methods in an interface are abstract, and there are no interface class methods.
\end{itemize}
Those rules can be applied on any given \mL~individually and in full isolation.

### 2017-05-29 FACS/section4.tex
Local changes

\saveSpace\saveSpace\section{Improving Reuse}\saveSpace
\name allows reuse even when subtyping is impossible.
In \name traits do not induce a new (externally visible) type.
However, locally in the trait the programmer can use the special self-type \Q@This@ in order to denote the 
type of \Q@this@.
That is: the program is agnostic to what the \Q@This@ type is going to be, so it can
be later assigned to any (or many) classes. 
The idea is that during flattening, \Q@This@ will be replaced with the actual class name.
In this way, \name allows reuse even when subtyping is
impossible, that is, when a method parameter has type \Q@This@ (also called \emph{binary
  methods}~\cite{bruce96binary}). 
This type of situations is the primary motivator
for previous work aiming at separating inheritance from subtyping.
Leveraging on the \Q@This@ type, we can also provide self-instantiation (trait methods can create instances of the class using them) and smootly integrate state and traits: a challenging problem that has limited the flexibility of traits and
reuse in the past.
%First we show how self-instantiation and state are already avaible in the simplest trait model,
%then we show how to radically improve the

@@

To illustrate how \name improves reuse, we will show a novel approach
to deal with \emph{state} in traits.  The idea of summing pieces of
code is very elegant, and has proven very successfull in module
composition languages~\cite{ancona2002calculus} and several trait
models~\cite{Traits:ECOOP2003,Bergel2007,BETTINI2013521,DBLP:journals/iandc/LagorioSZ12}.  However the research
community is struggling to make it work with object state (constructor
and fields) while achieving the following goals:

%\begin{itemize}
%complicated discussions on this point \item keep sum associative and commutative,
%\item
(1) managing fields in a way that borrows the elegance of summing methods;
%\item
(2) actually initialize objects, leaving no null fields;
%\item
(3) making it easy to add new fields;
%\item
(4) allowing a class to create instances of itself.
%\end{itemize}

In the related work we will show some alternative ways to handle
state.  However the purest solution just requires methods. The idea is
that the trait code just uses getter/setters/factories, while leaving
to classes the role to finally define the fields/constructors. That
is, the class has syntax richer that the trait one, allowing
declaration for fields and constructors.  This approach is very
powerful as illustrated by Wang et al.~\cite{wang2016classless}.

\paragraph{Immutable Points} Consider, for example, two simple 
traits that deal with \emph{immutable point} objects. That is, points
in the cartesian plane (with coordinates \lstinline{x} and
\lstinline{y}). The first trait provides a \emph{binary method} that 
sums the point object with another point to return a new point. 
The second trait provides a similar operation that does multiplication 
instead.
\saveSpace 
\begin{lstlisting}
  pointSum: { method int x()  method int y()//getters
    class method This of(int x,int y)//factory method
    method This sum(This that)
      This.of(this.x()+that.x(),this.y()+that.y())//self instantiation
    }
  pointMul: { method int x() method int y()//repeating getters
    class method This of(int x,int y)//repeating factory
    method This mul(This that)
      This.of(this.x()*that.x(),this.y()*that.y())
    }
\end{lstlisting}
\saveSpace
\noindent As we can see, all the state operations (the getters for the 
\lstinline{x} and \lstinline{y} coordinates) are represented as {\bf abstract} methods.
Notice the abstract \Q@class method This of(..)@ which acts as a constructor
for points:
A class method is similar to a \Q@static@ method in Java but can be abstract. 
As for instance methods, they are late bound:  flattening can provide an implementation for them.
abstract class method are very similar to the original concept of member functions in the module composition setting.

\subsection{A First Attempt at Composition}
According to the general ideas about trait composition presented in
Section~\ref{sec:separate}, a first attempt at composing the two traits providing
two different operations on points is:
\saveSpace\begin{lstlisting}
  Point:Use pointSum,pointMul
\end{lstlisting}  \saveSpace

\noindent However this fails since methods \Q@x@,\Q@y@ and \Q@of@ are still abstract!
Instead a user could write something similar to:

@@

\subsection{The \name approach to State}

@@

\paragraph{Coherent classes} A more detailed definition of coherent
classes is given next:
\begin{itemize}
\item a class with no abstract methods is coherent (just like Java
  \Q@Math@, for example). Such classes are useful for calling class/static methods.
\item a class with a single abstract \Q@class@ method returning \Q@This@
is coherent if all the other abstract methods can be seen as \emph{abstract state
operations} over one of its argument.
For example,
if there is a \Q@class method This of(int x, int y)@ as before,
then
\begin{itemize}
\item a method \Q@int x()@ is interpreted as an abstract state method: a \emph{getter} for \Q@x@.
\item a method \Q@Void x(int that)@ is a \emph{setter} for x.
\end{itemize}
\end{itemize}
\noindent
While getters and setters are fundamental operations, we can immagine
more operations to be supported; for example:
\begin{itemize}
\item \Q@method This withX(int that)@
may be a ``wither", doing a functional field update: it creates a new instance that is like \Q@this@ but where field \Q@x@ has now \Q@that@ value.
\item \Q@method Void update(int x,int y)@
may do two field update at a time.
\item\Q@method This clone()@ may do a shallow clone of the object.
\end{itemize}

@@

\paragraph{Points in \name:}
In \name and with our approach to handle the state, 
\lstinline{pointSum} and \lstinline{pointMul} can be directly composed:
\saveSpace
\begin{lstlisting}
  PointAlgebra:Use pointSum,pointMul
\end{lstlisting}  
\saveSpace
\noindent
  Note how we can declare the methods independently and compose the result
  as we wish.

  \paragraph{Improved solution} In this first attempt, we repeated the abstract methods \Q@x@,\Q@y@ and \Q@of@.
  Moreover, in addition to \Q@sum@ and \Q@mul@ we may want many
  operations over points. It is possible to improve reuse
  and not repeat such abstract definitions by abstracting the common
  abstract definitions into a trait \Q@p@: 
\saveSpace
\begin{lstlisting}
  p: { method int x() method int y()
    class method This of(int x,int y)
    }
  pointSum:Use p, { method This sum(This that)
      This.of(this.x()+that.x(),this.y()+that.y())
    }
  pointMul:Use p, { method This mul(This that)
      This.of(this.x()*that.x(),this.y()*that.y())
    }
  pointDiv: ...
  PointAlgebra:Use pointSum,pointMul,pointDiv,...
\end{lstlisting}
\saveSpace      
Now the code is fully modularized, and each trait handles exactly one method.

\subsection{State Extensibility}
Programmers may want to extend points with more state. For example 
they may want to add colors to the points. A first attempt at doing
this would be:
\saveSpace
\begin{lstlisting}
  colored:{ method Color color() }
  Point:Use pointSum,colored //Failure: resulting class not coherent
\end{lstlisting}
\saveSpace
This first attempt does not work: the abstract color method
is not a getter for any of the parameters of 
\Q@ class method This of(int x,int y)@. 
A solution is to provide a richer factory:
\saveSpace
\begin{lstlisting}
  CPoint:Use pointSum,colored,{
    class method This of(int x,int y) This.of(x,y,Color.of(/*red*/))
    class method This of(int x, int y,Color color)
    }
\end{lstlisting}
\saveSpace
\noindent 
where we assume support for overloading on different number of parameters.
This is a reasonable solution, however the method \Q@CPoint.sum@ resets
the color to red: we call the \Q@of(int,int)@ method, that now
delegates to \Q@of(int,int,Color)@ by passing red as the default field
value.  What should be the behaviour in this case?  If our abstract
state supports withers, instead of writing \Q@This.of(...)@ we can use
\Q@this.withX(newX).withY(newY)@ in order to preserve the color from
\Q@this@.  This solution is still not satisfactory: this design ignores
the color from \Q@that@.

>>>>

\saveSpace\saveSpace\section{Improving Reuse}\saveSpace
\name allows reuse even when subtyping is impossible.
In \name traits do not induce a new (externally visible) type.
However, locally in the trait the programmer can use the special self-type \Q@This@ in order to denote the 
type of \Q@this@.
That is: the program is agnostic to what the \Q@This@ type is going to be, so it can
be later assigned to any (or many) classes. 
The idea is that during flattening, \Q@This@ will be replaced with the actual class name.
In this way, \name allows reuse even when subtyping is
impossible, that is, when a method parameter has type \Q@This@ (also
called a \emph{binary
  method}~\cite{bruce96binary}). 
This type of situations is the primary motivator
for previous work aiming at separating inheritance from subtyping.
Leveraging on the \Q@This@ type, we can also provide self-instantiation (trait methods can create instances of the class using them) and smoothly integrate state and traits: a challenging problem that has limited the flexibility of traits and
reuse in the past.
%First we show how self-instantiation and state are already avaible in the simplest trait model,
%then we show how to radically improve the

@@

To illustrate how \name improves reuse, we will show a novel approach
to deal with \emph{state} in traits.  The idea of summing pieces of
code is very elegant, and has proven very successfull in module
composition languages~\cite{ancona2002calculus} and several trait
models~\cite{Traits:ECOOP2003,Bergel2007,BETTINI2013521,fjig}.  However the research
community is struggling to make it work with object state (constructors
and fields) while achieving the following goals:

\begin{itemize}
%complicated discussions on this point \item keep sum associative and commutative,
\item managing fields in a way that borrows the elegance of summing methods;
\item actually initialize objects, leaving no null fields;
\item making it easy to add new fields;
\item allowing a class to create instances of itself.
\end{itemize}

In the related work we will show some alternative ways to handle
state.  However the purest solution requires methods only. The idea is
that the trait code just uses getter/setters/factories, while leaving
to classes the role to finally define the fields/constructors. That
is, classes have syntax richer than traits, allowing
declaration for fields and constructors.  This approach is very
powerful as illustrated by Wang et al.~\cite{wang2016classless}.

\paragraph{Modeling Points} Consider, for example, two simple 
traits that deal with \emph{point} objects. That is, points
in the cartesian plane (with coordinates \lstinline{x} and
\lstinline{y}). The first trait provides a \emph{binary method} that 
sums the point object with another point to return a new point. 
The second trait provides a similar operation that does multiplication 
instead.
\saveSpace 
\begin{lstlisting}
  pointSum: { method int x()  method int y()//getters
    class method This of(int x,int y)//factory method
    method This sum(This that)
      This.of(this.x()+that.x(),this.y()+that.y())//self instantiation
    }
  pointMul: { method int x() method int y()//repeating getters
    class method This of(int x,int y)//repeating factory
    method This mul(This that)
      This.of(this.x()*that.x(),this.y()*that.y())
    }
\end{lstlisting}
\saveSpace
\noindent As we can see, all the state operations (the getters for the 
\lstinline{x} and \lstinline{y} coordinates) are represented as {\bf abstract} methods.
Notice the abstract \Q@class method This of(..)@ which acts as a constructor
for points:
a class method is similar to a \Q@static@ method in Java but can be abstract. 
As for instance methods, they are late bound:  flattening can provide an implementation for them.
Abstract class methods are very similar to the original concept of member functions in the module composition setting.

\subsection{A First Attempt at Composition with a Traditional Model of
Traits}
According to the general ideas about trait composition presented in
Section~\ref{sec:separate}, a first attempt at composing the two traits providing
two different operations on points is:
\saveSpace\begin{lstlisting}
  Point:Use pointSum,pointMul
\end{lstlisting}  \saveSpace

\noindent However \emph{in a traditional trait model} this would fail since methods \Q@x@,\Q@y@ and \Q@of@ are still abstract!
Instead a user could write something similar to:

@@

\subsection{The \name Approach to State: Coherent Classes}

@@

\paragraph{Coherent classes} A more detailed definition of coherent
classes is given next:
\begin{itemize}
\item a class with no abstract methods is coherent (just like Java
  \Q@Math@, for example). Such classes are useful for calling class/static methods.
\item a class with a single abstract \Q@class@ method returning \Q@This@
is coherent if all the other abstract methods can be seen as \emph{abstract state
operations} over one of its argument.
For example,
if there is a \Q@class method This of(int x, int y)@ as before,
then
\begin{itemize}
\item a method \Q@int x()@ is interpreted as an abstract state method: a \emph{getter} for \Q@x@.
\item a method \Q@Void x(int that)@ is a \emph{setter} for x.
\end{itemize}
\end{itemize}
\noindent
While getters and setters are fundamental operations, it is possible to
support more operations. For example:
\begin{itemize}
\item \Q@method This withX(int that)@
may be a ``wither", doing a functional field update: it creates a new instance that is like \Q@this@ but where field \Q@x@ has now \Q@that@ value.
\item \Q@method Void update(int x,int y)@
may do two field updates at a time.
\item\Q@method This clone()@ may do a shallow clone of the object.
\end{itemize}

@@

\paragraph{Points in \name:}
In \name and with our approach to handle the state, 
\lstinline{pointSum} and \lstinline{pointMul} can indeed be directly composed:
\saveSpace
\begin{lstlisting}
  PointAlgebra:Use pointSum,pointMul //coherent class completion!
\end{lstlisting}  
\saveSpace
\noindent
  Note how we can declare the methods independently and compose the result
  as we wish.

  \paragraph{Improved solution} So far the current solution still
  repeats the abstract methods \Q@x@, \Q@y@ and \Q@of@.
  Moreover, in addition to \Q@sum@ and \Q@mul@ we may want many
  operations over points. It is possible to improve reuse
  and not repeat such abstract definitions by abstracting the common
  abstract definitions into a trait \Q@p@: 
\saveSpace
\begin{lstlisting}
  p: { method int x() method int y()
    class method This of(int x,int y)
    }
  pointSum:Use p, { method This sum(This that)
      This.of(this.x()+that.x(),this.y()+that.y())
    }
  pointMul:Use p, { method This mul(This that)
      This.of(this.x()*that.x(),this.y()*that.y())
    }
  pointDiv: ...
  PointAlgebra:Use pointSum,pointMul,pointDiv,...
\end{lstlisting}
\saveSpace      
Now the code is fully modularized, and each trait handles exactly one method.

\subsection{State Extensibility}
Programmers may want to extend points with more state. For example 
they may want to add colors to the points. A first attempt at doing
this would be:
\saveSpace
\begin{lstlisting}
  colored:{ method Color color() }
  Point:Use pointSum,colored //Fails: class not coherent
\end{lstlisting}
\saveSpace
This first attempt does not work: the abstract color method
is not a getter for any of the parameters of 
\Q@ class method This of(int x,int y)@. 
A solution is to provide a richer factory:
\saveSpace
\begin{lstlisting}
  CPoint:Use pointSum,colored,{
    class method This of(int x,int y) This.of(x,y,Color.of(/*red*/))
    class method This of(int x, int y,Color color)
    }
\end{lstlisting}
\saveSpace
\noindent 
where we assume support for overloading on different number of parameters.
This is a reasonable solution, however the method \Q@CPoint.sum@ resets
the color to red: we call the \Q@of(int,int)@ method, that now
delegates to \Q@of(int,int,Color)@ by passing red as the default field
value.  What should be the behaviour in this case?  If our abstract
state supports withers, we can use
\Q@this.withX(newX).withY(newY)@, instead of writing \Q@This.of(...)@, in order to preserve the color from
\Q@this@.  This solution is still not satisfactory: this design ignores
the color from \Q@that@.
